---
title: "Probability"
subtitle: "Fundamentals of Data Science"
author: "Jeremy Teitelbaum"
format: html
title-block-style: plain
---

# Populations and Samples

## Probability Theory

Probability theory is based on:

- An underlying collection $S$ of all possible outcomes  (a *population* or *sample space*) of an experiment. 
- A rule $P$ that assigns a number between zero and one to each subset of the sample space satisfying certain rules. 

## Sample Space


For example:

- For a flip of a single coin, the possible outcomes are Heads and Tails and the sample
space has two elements. For multiple flips, the outcomes are sequences of Heads and Tails.

- For a measurement of temperature, we might model the possible outcomes, or the sample space,   as all real numbers, recognizing that only some of them are actually possible results of the experiment.


## Simple Events

The elements of the sample space or population are the *outcomes* or *simple events* or *sample points*. 

- For a flip of a coin, the possible outcomes are Heads or Tails.  For multiple flips,
the possible outcomes are particular sequences of Heads or Tails.

- For a measurement of temperature, a simple event would be a particular number obtained
at a particular time. 

## Events

Subsets of the population make up events or outcomes.

- Among the population made up of sequences of 10 coin flips, the subset consisting of
sequences containing at least 3 heads is an event. 

- Among the measurements of temperature, a measurement lying between say 22 and 25 degrees celsius would be an event. 

## Probability Measure

The last element of probability theory is the function P that assigns a number between 0 and 1 to every event such that

- $P(\emptyset)=0$
- $P(S)=1$. 
- If $A\cap B=\emptyset$ then $P(A\cup B)=P(A)+P(B)$.  This is also required to hold for *infinite* collections of disjoint sets but we won't worry much about the foundations of probability. 

## Random Variables

A *random variable* is a rule that assigns a number to an event. 

- We can assign the value 1 to heads and 0 to tails.  This is a *bernoulli* random variable.
- Our sample space can be sets of 10 coin flips.  The number of heads is a random variable. 
- The measurement of temperature yields a number.  
- If we pick a person at random, we can assign the value 1 if they wear glasses and 0 if not.

## Discrete vs Continuous Random Variables

A discrete random variables takes "separate" values depending on the event.
A continuous random variable takes values in a range.

- Bernoulli random variable is discrete (0/1)
- Number of heads in 10 flips is discrete (takes values 0,...,10)
- Temperature is continuous (in principle can get any reading)
- Mass of a penguin is continuous
- Species of penguin is discrete


## Events and Random Variables

Specifying a value, or a range of values, for a random variable defines an event.

## Bernoulli example

- Sample space is $\{H,T\}$
- $P(H)=p$
- $X$ is the random variable with $X(H)=1$ and $X(T)=0$

Then:

- $X=1$ is the same as the event $H$
- $P(X=1)$=p

## Binomial Example

A binomial random variable (with parameters $n$ and $p$) is the sum of $n$
bernoulli random variables with probability $p$.  It corresponds
to flipping a coin (with $P(H)=p$) $n$ times and counting up the heads.

The probability of getting $k$ heads is
$$
P(k)=\binom{n}{k}p^k(1-p)^{n-k}
$$

```{r}
library(ggplot2)
x <- dbinom(seq(0, 10), 10, .3)
ggplot() +
    geom_bar(aes(x = seq(0, 10), y = x), stat = "identity") +
    scale_x_continuous(breaks = seq(0, 10)) +
    xlab("Number of Heads") +
    ylab("Probability") +
    ggtitle("10 Flips, P(H)=.3")
```


## Continuous example

- Sample space is the possible temperatures at a particular point in space and time.
- Random variable $T$ is a measure of temperature.
- $P(21<T<22)$ is the probability that the temperature is between 21 and 22 degrees.

## Probability density functions

In the continuous case, probability is measured by a probability density function $P(x)$. The classic example is the normal (bell-shaped) curve.


## Area gives probability

```{r echo=FALSE, output=FALSE}
library(tidyverse)
```
```{r echo=FALSE}
temperature <- seq(21, 23, .001)
density <- dnorm(temperature, 22, .2)
p <- pnorm(22.3, 22, .2) - pnorm(21.7, 22, .2)
ggplot() +
    geom_line(aes(x = temperature, y = density)) +
    geom_ribbon(aes(x = temperature, ymin = 0, ymax = if_else(temperature >= 21.7 & temperature <= 22.3, density, 0)), fill = "gray")
```

If $P(x)$ is the density function, then:

> the probability that $x$ lies between $a$ and $b$ is the area under density function between $a$ and $b$.

The shaded area gives probability `r format(p,digits=2,nsmall=2)` for temp between 21.7 and 22.3.


## Standard Normal

A normal curve is defined by two parameters:

- the mean $\mu$, which sets the *location*
- the standard deviation $\sigma$ or its square, the variance $\sigma^2$, which sets the *scale*.

If $x$ is a normal random variable with mean $\mu$ and variance $\sigma^2$, then
$$
z = \frac{x-\mu}{\sigma}
$$
is a normal random variable with mean $0$ and variance $1$.  This is called a $z$-score or a *standard normal* variable. 

## Cumulative Distribution

The cumulative distribution is a function $f(x)$ such that $f(x)$ is the the percentage of samples that  are less than $x$.  

```{r echo=FALSE}
samples <- rnorm(100, 22, .2)
true_cdf <- pnorm(temperature, 22, .2)
ggplot() +
    stat_ecdf(aes(x = samples, color = "Empirical CDF")) +
    geom_line(aes(x = temperature, y = true_cdf, color = "True CDF")) +
    scale_color_manual(values = c("Empirical CDF" = "black", "True CDF" = "red")) +
    geom_hline(yintercept = .5) +
    geom_vline(xintercept = median(samples))
```

So the median of the samples occurs where the $y$-axis is $.5$.

## Quantiles

If $q$ is between $0$ and $1$, then the $q^{th}$ quantile $Q$ of a random variable $x$ is the value of
$x$ such that the fraction of the population with $x<Q$
is $q$.

The median of $x$ is the $.5$ quantile for $x$ because half of the population has values less than the median.

One can read quantiles from the cumulative distribution.

## Order Statistics

The sample median and the sample quantiles (such as the 25th percentile or 75th percentile) are examples of *order statistics.*

The smallest element, the second smallest element, and so on are other examples of *order statistics*.

## Order Statistics example

We take 100 samples from a normal distribution and compute the median, minimum, and maximum.  Then we do that 10000 times and produce a histogram. 

```{r echo=FALSE}
minima <- replicate(10000, min(rnorm(100, 0, 1)))
maxima <- replicate(10000, max(rnorm(100, 0, 1)))
medians <- replicate(10000, median(rnorm(100, 0, 1)))
ggplot() +
    geom_histogram(aes(x = minima, y = after_stat(density), color = "Minimum"), bins = 100) +
    geom_histogram(aes(x = maxima, y = after_stat(density), color = "Maximum"), bins = 100) +
    geom_histogram(aes(x = medians, y = after_stat(density), color = "Medians"), bins = 100) +
    scale_color_manual(values = c("Minimum" = "red", "Maximum" = "green", "Medians" = "blue"))
```


## The multinomial distribution

The multinomial distribution arises when you have $n$ outcomes for your experiment, say $x_1,\ldots, x_n$;
and the probability of getting $x_i$ is $p_i$.  Here
we have to have
$$
\sum p_{i}=1.
$$

This generalizes the bernoulli distribution.

## Mean 
The mean  of a random variable is perhaps the most important statistic
associated with a probability space.

The mean is the "average value" of the random variable. 

## Mean -- discrete case

In the discrete case:

$$
\overline{x}=\sum_{a\in X} x(a)p(a)
$$

In other words, the mean of $\overline{x}$ is the sum of $x$ at each event, weighted by the probability of that event.

## Mean of a bernoulli random variable

If $x$ is bernoulli, with $p(x=1)=p$, then the mean of $x$ is
$$
p(1)+(1-p)(0)=p.
$$

## Mean of a binomial random variable

If $x$ is binomial, corresponding to the sum of $N$ bernoulli random variables
with probability $p$, then
$$
\overline{x} = \sum_{0\le i\le N} i\binom{N}{i}p^{i}(1-p)^{N-i}=Np
$$

## Mean -- continuous case

The mean of a continuous random variable is given by an integral:

$$
\overline{x} = \int_{X} xp(x) dx
$$

where $p(x)$ is the probability density.  This is the limiting case of the formula in the discrete case.



## Mean of standard normal

The mean of the standard normal is zero. 

## Variance and standard deviation

The variance of a random variable measures how it is distributed around its mean value.

In the case of a discrete random variable with outcomes values $x$ having probability $p(x)$, the variance is
$$
\sum_{x} (x-\overline{x})^2p(x)
$$


This is the same as 

$$
\sigma^2 = \overline{x^2}-(\overline{x})^2
$$

The standard deviation is the square root of the variance.

## Variance of Bernoulli

In the Bernoulli case, the variance is 
$$
(1-p)^2p+p^2(1-p)=p(1-p).
$$

Notice that the maximum variance happens when $p=1/2$.  

## Variance of Binomial

A binomial random variable with probability $p$ and $n$ trials is a sum of $n$ bernoulli random variables with probability $p$.  Using the formula
you get
$$
\sigma^2 = \sum_{i=0}^{n}(i-np)^2\binom{n}{i}p^{i}(1-p)^{n-i}
$$

This turns out to be
$$
\sigma^2 = np(1-p).
$$

## Hypothesis Testing

A *statistical hypothesis* is a claim about a particular population. A *hypothesis test* is a method to determine which of two contradictory hypotheses is supported by the data.

Underlying idea: a lot of surprising things happen by chance.  If you do an experiment and observe an effect, that might be the result of pure chance. Can
you quantify that?

## An example

Suppose we have a coin and we'd like to do some testing to determine if we 
have reason to suspect that the coin is biased.  Put another way, you'd like
to know if this coin behaves differently from a reference, standard coin that is fair. 

Note this is more common than you might think. It might arise in the following
circumstance in "real life." You have two web pages, your current one and a proposed new one.  You'd like to know if seeing the proposed one increases the chance of a viewer clicking through to something.  This is called A/B testing and it amounts
to comparing the probability of click-through in the reference case to the proposed case. 

## Null (and alternative) hypotheses

The Null hypothesis is the hypotheses that our coin is fair, or that our two web pages yield the same results, or more generally that the observations we make
are accounted for only by chance and not by some underlying effect. So our null hypothesis for our coin is "P=.5".

An alternative hypothesis is a statement that contradicts the null hypothesis.  For example, "P>.5" or "P<.5" or "P is different from .5."

## Test statistic

A test statistic is a measurement of the data used to draw conclusions about the sample.

## Back to our example

For our test statistic, we are going to use the fraction of times we get a head
in N flips.

In the A/B testing situation, our test statistic would be the fraction of times a person "clicked through" when given the proposed web site.

Intuitively, if the fraction of heads differs significantly from the expected fraction of heads (.5) then we take that as strong evidence for the unfairness of our coin (or the increased value of our proposed web page).

How can we quantify this?

## Rejection region and "significance"

To make things concrete, suppose the coin is fair (in other words, the null hypothesis is true) we flip the coin $10$ times.  If the coin is far, we expect to get
roughly 5 heads.  

There's a long tradition of saying something unlikely is "significant" if the
chance of it occurring, assuming the null hypothesis, is less than .05 or one in twenty.  

The chance of getting $0$, $1$, $9$, or $10$ heads is $.02$.  If we allow
$2$ or $8$ heads in addition, the chance is about $11%$, so if we set our significance level at $.05$ we *reject the null hypothesis* if our experiment yields $0$, $1$, $9$,
or $10$ heads.

This is the "rejection region."
```{r}
library(ggplot2)
rejection2 <- function(n,d=10) {
    results <- data.frame(x=seq(0,n),y=dbinom(seq(0,n),n,.5),keep=sapply(seq(0,n),function(x) (x<qbinom(.025,n,.5)) | (x>qbinom(.975,n,.5))))
    ggplot(data=results,aes(x=x,y=y,fill=keep))+geom_bar(stat="identity")+
    scale_x_continuous(breaks=seq(0,n,d))+ggtitle("Two sided rejection region at alpha=.05")

}
rejection1 <-function(n,d=10) {
    results <- data.frame(x=seq(0,n),y=dbinom(seq(0,n),n,.5),keep=sapply(seq(0,n),function(x) (x>qbinom(.95,n,.5))))
    ggplot(data=results,aes(x=x,y=y,fill=keep))+geom_bar(stat="identity")+
    scale_x_continuous(breaks=seq(0,n,d))+ggtitle("One sided rejection region at alpha=.05")
}
```

```{r}
rejection2(10, 1)

```

## One-sided test

Suppose you want evidence that your coin is *more likely* to get heads. So your null hypothesis is that your coin has $p=.5$.  Your alternative hypothesis is $p>.5$.

The probability of getting 0,1,2 heads is $.054$, which is a bit larger than
$.05$.  So the one-sided test would reach significance only at 0 or 1 heads
same as the two-sided test.  But with more samples, the "one-sided" rejection
region would be wider on the left than the two-sided one.

But suppose we did 50 flips? Then the one- and two-sided limits are slightly different. 


```{r}
rejection2(50,1)
```
```{r}
rejection1(50, 1)
```

## Normal approximation

For large $n$, the binomial distribution distribution with probability $p$ becomes
a version of the normal distribution with mean $Np$ and standard deviation $\sqrt{Np(1-p)}$.

```{r}
compare <- function(n, p = .5) {
    results <- data.frame(x = seq(0, n), y = dbinom(seq(0, n), n, p), z = dnorm(seq(0, n), mean = p * n, sd = sqrt(n * p * (1 - p))))
    ggplot(data = results) +
        geom_bar(aes(x = x, y = y), stat = "identity", fill = "lightgreen") +
        geom_line(aes(x = x, y = z), color = "red", size = 2) +
        ggtitle(glue("Normal Approximation to the binomial with n={n} and p={p}"))
}
compare(100, .3)
```

So one can use the normal distribution to determine the rejection region if $n$ is large.

## Errors

Two things can go wrong:

Type I error: You reject the null hypothesis, but the null hypothesis is true.
The probability of a Type I error is something you choose when you set the significance level.  This is usually called $\alpha$.

Type II error: You accept the null hypothesis when it is false.  In this case, 
you've *missed* an actual effect. This is called $\beta$. Typically, if you
make $\alpha$ smaller (thus reducing the chance of a Type I error) you make
$\beta$ bigger (increasing the chance of a type II error). 

## Statistical Power

Informally, statistical power measures the ability of an experiment to detect a real effect.  If a study has high power, then you are very unlikely to make a Type II error.

For example, return to the coin flipping problem (or the A/B testing problem).

Suppose we flip our coin $20$ times and our null hypothesis is that $p=.5$ 
If our significance level is $.05$, we will reject the null hypothesis and conclude
that the coin is *not* fair (and biased towards heads) provided we get $15$ or more  heads.  

```{r}
# look at the probability density for this case
probs <- dbinom(seq(0, 20), 20, .5)
# The qbinom function tells us the threshold
rejection <- qbinom(.95, 20, .5)
# We check this by comparing the probability of $15-20$ vs $14-20$ heads:
sum(probs[16:21]) # remember probs[i] is the chance of i-1 heads
sum(probs[15:21])
# The chance of $15-20$ heads is greater than $.05$.
```

Now suppose the coin is *not* fair and $p=.6$.  What is the chance that we *accept* the null hypothesis and conclude, falsely, that the coin is fair?  It is the chance
that we get $14$ or fewer heads when $p=.6$.

```{r}
probs6 <- dbinom(seq(0, 20), 20, .6)
sum(probs6[1:15])
```

This is $87$ percent!  In other words, our experiment is very *unlikely* to detect
the unfairness of the coin if the unfairness is only the difference between $p=.6$ and
$p=.5$

But if the coin is *very* unfair, with, say $p=.8$, then we find:

```{r}
probs8 <- dbinom(seq(0, 20), 20, .8)
sum(probs8[1:15])
```

We have only a 20% chance of a Type II error so there's an 80% chance we'll detect the problem.

Now suppose we use $100$ flips.

```{r}
rejection <- qbinom(.95, 100, .5)
probs6 <- dbinom(seq(0, 100), 100, .6)
sum(probs6[1:rejection])
```

Now we have a 70% chance of detecting the difference unfairness of $p=.6$.
<!--
some things to talk about:
fisher exact test  (likelihood of results at random)
datatypes and types of random variables (ints are discrete/continuous?)
relationship between *density function* and *histogram*
empirical distribution
-->


