---
title: "R Notebook Walkthrough"
output:
  html_document:
    df_print: paged
subtitle: "Fundamentals of Data Science"
author: "Jeremy Teitelbaum"

---

Start out with a code cell saying "Hello World"

```{r}
print("Hello World")
```
The `cat` command is actually probably more useful than print:
```{r}
cat("hello world")
```
## Variables, Types, and Assignment

In R, the assignment operator is <-, not =.  This takes some getting used to.

```{r}
count <- 5
name <- "Jeremy Teitelbaum" # string types are called chr for character
paragraph <- "Far across the misty mountains cold,
to dungeons deep and caverns cold,
we must away,
ere break of day
to seek our long forgotten gold."
pi <- 3.14159 # R doesn't use integer types unless you force it to, numbers are "num" # nolint: line_length_linter.
epsilon <- 1e-6
count <- 5L # this forces an integer
students <- c("Jeremy", "Phillip", "Sara", "Molly")
hot_dog <- TRUE # note all caps unlike Python; false is FALSE
```
In R, you can give names to the elements of a vector.
```{r}
print("hello")
```

```{r}
names(students) <- c("President", "Vice President", "Treasurer", "Secretary")
print(names(students))
print(students["President"])
print(students)
```



The cat command is a print command that "concatenates" its arguments; it needs an *explicit* newline.
```{r}
print(students)
print(count)
cat("Students:", students, "\n")
print(epsilon)
cat("The value of epsilon is:", epsilon, "\n")
print(paragraph)
cat(paragraph)
```
The [1] at the beginning of each of these things reflects the fact that in R *everything is a vector*.  So it is telling you that the first thing there is element 1 of the vector.

The c() command makes a vector of its arguments.  It forces everything to be of the same type.

```{r}
str_list <- c("Jeremy", 25, 1.34, FALSE) # everything becomes a string
int_list <- c(1, 2, 3, 4, 5)
float_list <- c(1, 2, 3.5, 4)
```
## Arithmetic

R does all arithmetic on vectors/lists.  It one is shorter than the other, it repeats the shorter one, but the length of the longer has to be a multiple of the shorter.

```{r}
a <- 1
b <- 2
a + b
```

```{r}
a <- c(1, 2, 3, 4, 5)
b <- 4
a + b
```

```{r}
a <- c(1, 2, 3, 4, 5, 6)
b <- c(10, 11)
a + b
```

```{r}
a <- c(1, 2, 3, 4, 5)
b <- c(1, 2)
a + b
```

```{r}
a / 5
```

```{r}
# integer division (// in python)
a <- 5L
b <- 3
a %/% b
```

```{r}
# remainder (% in python)
a <- 5
b <- 3
a %% b
```

```{r}
a <- c(1, 2, 3, 4, 5)
a^2
```

```{r}
print(a^2 == a)
print(a^2 > a)
print(a^2 == 4)
```
## Operations on strings and lists

```{r}
first_name <- "Jeremy"
last_name <- "Teitelbaum"
nchar(first_name)
```

```{r}
paste(first_name, last_name) # spaces by default
```

```{r}
paste(first_name, last_name, sep = "") # no space
```

```{r}
paste(c(1, 2, 3), "Jeremy") # remember functions work across vectors
```
## Substrings

In R, **you always count from 1** (big difference from python)
```{r}
first_name[1] # another difference from Python
```
```{r}
a <- substr("Jeremy", 1, 1)
b <- substr("Jeremy", 1, 3)
cat(a, b, paste(a, b, sep = ""))
```
## Slicing lists

```{r}
nums <- 0:10 # generates a sequence from 0 to 10 INCLUSIVE (compare python)
print(nums)
print(nums[c(1, 3)]) # you can pass a list of indices to a subscript
```
```{r}
sqrs <- nums^2
sqrs[seq(1, 10, 2)]
```
In R, negative numbers in seq mean "omit" so this means omit entries 2 through 5. You can't mix positive and negative numbers
```{r}
rev <- nums[seq(-2, -5)]
print(rev)
```
```{r}
rev(nums) # reverses the list
```

## Libraries and packages

Use the Rstudio package manager to add libraries to your installation, but to use them you need to use the library function.
The tidyverse library is something we will use a lot.

```{r}
library(tidyverse)
```
```{r}

```

## Plotting
```{r}
library(ggplot2)
```


```{r}
x <- seq(-10, 10, .1)
y <- x**2
data <- tibble("x" = x, "y" = y)
```

```{r}
ggplot(data = data, aes(x = x)) +
    geom_point(aes(y = y), color = "red") +
    ggtitle("A Parabola") +
    scale_x_continuous(breaks = seq(-10, 10, 1)) +
    scale_y_continuous(breaks = seq(0, 100, 20))
```



```{r}
x <- seq(-10, 10, .1)
y <- cos(x)
data <- tibble("x" = x, "y" = y)
ggplot(data = data, aes(x = x)) +
    geom_line(aes(y = y), color = "darkgreen") +
    ggtitle("A Cosine Curve") +
    scale_x_continuous(breaks = seq(-10, 10, 1)) +
    scale_y_continuous(breaks = seq(-1, 1, 5))
```

